Supplementary methods

### 2020-08-31 text/survival.Rmd
final edits Dennis

To illustrate the model, we simulate survival time curves
for four related groups (**figure 5A**. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated accurately without the confounding
effect of age.

@@

**Figure 5: Comparison of different cancers, simulated and actual results
results.** To illustrate the risk model, we compare survival of four simulated
cohorts. A) A group consisting of randomly selected 30 year old women (light
blue) vs. an equal sized group of 30 year old women with cancer, k = 2 (light
orange). To illustrate how the mix of ages alters the survival curve, we compare
a randomly selected group of people with ages uniformly distributed from 30 to
70 (dark blue) and a randomly selected group with uniformly distributed ages and
cancer k = 2 (dark orange). B,C,D,E) We compare a randomly selected group with
correct age distribution for the given cancer (turqoise) to the model predicted
survival for the given average k (blue) and the actual survival for the patients
in our dataset (orange). Differences with actual data are likely due to a
mixture of cluster identities which we exclude from these simulations for
simplicity. F) Effects are compared between cancer type by showing the
probability density for the posterior value of the $k_{tis}$ multiplier. Larger
values correspond to decreased age-independent life expectancy.

@@

**Figure 6 Cluster specific $k_{cl}$ values for each cancer type.** The overall
$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific value
represents the relative aggressiveness of each cluster within the cancer type.
For example clusters 1, 2, and 6 Glioblastoma (GBM) are apparently much less
deadly than other types of GBM. A similar situation can be seen with Pancreatic
Cancer (PAAD) where 1, 6, and 8 are less deadly compared to clusters 7, 9, 10.
For some cancer types there are few data points and little ability to estimate
values with precision (DLBC and KICH for example).

>>>>

To illustrate the model, we simulate survival time curves
for four related groups (**figure 5A**). We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated accurately without the confounding
effect of age.

@@

**Figure 5: Comparison of different cancers, simulated and actual results.** To
illustrate the risk model, we compare survival of four simulated cohorts. A) A
group consisting of randomly selected 30 year old women (light blue) vs. an
equal sized group of 30 year old women with cancer, $k = 2$ (light orange).
Compare a randomly selected group of people with ages uniformly distributed from
30 to 70 (dark blue) and a randomly selected group with uniformly distributed
ages and cancer $k = 2$ (dark orange). B,C,D,E) Compare random group with
correct age distribution for the given cancer (turqoise) to the model predicted
survival for the given average k (blue) and the actual survival for the patients
in our dataset (orange). Differences with actual data are likely due to a
mixture of cluster identities which we exclude from these simulations for
simplicity. F) Effects are compared between cancer type by showing the
probability density for the posterior value of the $k_{tis}$ multiplier. Larger
values correspond to decreased age-independent life expectancy.

@@

**Figure 6 Cluster specific $k_{cl}$ values for each cancer type.** The overall
$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific value
represents the relative aggressiveness of each cluster within the cancer type.
Alpha transparency (α) is set to reflect varying confidence in the posterior
distribution when the number of observations is N<10, α = 0.1; N<25, α = 0.5; N
≥ 25 α = 1.0.
